---
title: "A decomposition approach for discrete time multistate models"
author: "Tim Riffe"
date: "10/11/2019"
output: pdf_document
bibliography: references.bib
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```
\[Edited ~20.12.2019\]

# Background
Multistate models are useful for calculating summary statics from dynamic processes defined on the basis of transitions, for example a healthy life expectancy. Examples of such models are found in the areas of health, family, migration, and labor demography among others. Often, we would like to explain differences between the sexes, over socioeconomic gradients, or changes over time within groups. Part of the lure of multistate models is the promise of a process-based model. Most often, the process itself is not analyzed, but rather it is assumed, estimated, and these results are used to calculate a desired static index-- a long run consequence of the model and its parameters. 

We would like to offer an approach to compare model results, by attributing differences in results to differences in their respective parameters. Multistate models have a functional form, which means that a given set of parameters $\theta$ (transition probabilities and possibly an initial state composition in this case) is used to deterministically calculate an aymptotic result, such as a state expectancy. Therefore differences between results can be attributed to elementwise differences between two versions of $\theta$. These contributions offer primary evidence of which forces account for differences observed, a first step for explaining differences, or a first clue for designing targetted interventions. 

There are three general decomposition approaches (that I know of) that can be used in settings such as this. These include the algorithm of stepwise replacement (ASR) [@andreev2002algorithm], the difference-scaled derivatives (DSD) approach [@caswell1989analysis], and pseudo-continuous decomposition (PCD) [@horiuchi2008]. The second approch DSD^[This method is primarily known as  a *lifetable response experiment* LTRE, but I opt to give it a more literal and descriptive name.] is the only one of these three approaches that has been used to decompose multistate models in the literature [@horvitz1997relative give an overview], but in principle these methods are interchangeable and they should give only slightly different results. I do not here scrutinize the choice of general decomposition method.

In this note I pose a previously undescribed conundrum in the decomposition of discrete time multistate models: When decomposing differences in this class of model, it matters how $f(\theta)$ is parameterized. It matters considerably, in ways that can potentially reverse an interpretation. To make matters worse, there are infinitely many parameterizations one can choose from. Each parameterization can be subject to a *valid* decomposition. Validity here means that the sum of the contributions from all parameters is equal to the difference in the objective index. Often this kind of validity is the only check given to the correctness of a decomposition, and that's why it is easy for this particular conundrum to go unnoticed.

The case at hand is in a sense undetermined, unsolvable. But I think one could choose, or narrow down, the parameterization design choices with an axiomatic approach. No one has stated what properties a parameterization should have in order to produce a plausible result. An understandable description of the nature of this problem requires some requisite knowledge, here given in brief. 

The core issue is that the probabilities of the $s+1$ transitions originating from a given state sum to one, and are therefore a composition, ergo constrained. Standard calculations of e.g. healthy life expectancy are based on only $s$ parameters per age and origin state. Decomposition is premised on hypothetical perturbations of parameters. Since parameters are constrained in this way, they cannot be considered independent, meaning that perturbation of one parameter necessarily affects the others. In the following I give a structured introduction to the problem at hand, on the basis of a toy example. 

# Toy example function

Let's assume a model with 3 ages and the states (`H`)ealthy, (`S`)ick and (`D`)ead. 

```{r, include=FALSE, message = FALSE, warning=FALSE}
library(here)
library(data.table)
library(tidyverse)
source(here::here("Code","R","Functions.R"))
TR    <- local(get(load(here::here("Data","Transitions","DCS","initdetail","TR_v06.Rdata"))))
TR    <- data.table(TR)
# TR: had to fix colnames problem
PREV  <- TR[ , get_prev_dt(.SD), by = list(sex, edu, time)]
# TR2 just means that there are now TWO living states,
TR2   <- collapseTR(TR = TR, PREV = PREV) %>% 
  filter(time == 2001, sex == "b", edu == "secondary")
TR2

data_2_U

```


# Acknowledgements
This exercise has benefitted from discussions with Christian Dudel, Kathryn Grace, Jutta Gampe, Marie-Pier Bergeron, James Oeppen, Alyson van Raalte, Hal Caswell, Jonas Schöley, Shiro Horiuchi, Daniel Schneider, Marilia Nepomuceno, Vera Pawlowsky-Glahn, and Juan José Egozcue.

# References
